---
title: "Tortoises and Iguanas"
title-block-banner: true
format:
  html:
    self-contained: true
    toc: true
    theme: yeti
  pdf: default
output:
  rmarkdown::html_document:
    theme: yeti
  pagedown::pdf_document: default
editor: visual
author:
  - "Charles Lehnen"
lan: en
---

```{r}

rm(list = ls())

```

```{r}

library(readr) 
library(dplyr)
library(ggplot2)
library(janitor)
library(stringr)
library(lubridate)
library(ggalluvial) # for the alluvial plot
library(webr) # for the stacked donut plot


```

```{r}

# Output folder
dir.create("../output/r_analysis", showWarnings = FALSE, recursive = TRUE)

```

Loading in our data

```{r}

# The working directory at render time differs depending on how/where this
# gets knitted, so try the candidate paths in order and use the first that exists.
read_first_existing <- function(candidates) {
  for (path in candidates) {
    if (file.exists(path)) {
      return(read.csv(file = path))
    }
  }
  stop("None of the candidate paths exist: ", paste(candidates, collapse = ", "))
}

dfclassifications <- read_first_existing(c(
  "../output/classifications.csv",
  "/output/classifications.csv",
  "output/classifications.csv"
))
dfclassifications_random <- read_first_existing(c(
  "../output/classifications_random_order.csv",
  "/output/classifications_random_order.csv",
  "output/classifications_random_order.csv"
))
dfmetadata <- read_first_existing(c(
  "../output/metadata/metadata.csv",
  "/output/metadata/metadata.csv",
  "output/metadata/metadata.csv"
))
```

explore data

```{r}

head(dfclassifications)

```

```{r}

summary(dfclassifications)

```

```{r}

str(dfclassifications)

```

Cleaning our data

```{r}
dfclassifications_clean <- dfclassifications %>%
  clean_names() %>%
  rename_with(~ "species", ends_with("species")) %>%
  filter(species != "") %>%
  # If a crop was labelled out of the buffered folder it carries a _buf100 suffix that the metadata doesn't have. drop it so the two match.
  mutate(crop_path = str_replace(image, "_buf\\d+(\\.[^.]+)$", "\\1")) %>%
  mutate(crop_path = str_replace(crop_path, "^\\./", ""))

dfmetadata_clean <- dfmetadata %>%
  clean_names() %>%

  # get column name formats to match
  mutate(crop_path = str_remove(cropped_image, "^.*output/cropped/")) %>%
  mutate(
    first_seg = str_replace(subfolder, "[/\\\\].*$", ""), 
    cam_id    = str_extract(first_seg, "\\d{3,}"),
    camera_trap_name = if_else(is.na(cam_id), "000", cam_id)
  ) %>%
  
  mutate(
    dt   = parse_date_time(date_time_original, orders = "Y:m:d H:M:S"),
    date = as_date(dt),
    hour = hour(dt)
  ) %>%

  # Santa Fe sits on the equator, so daylight is roughly 06:00-18:00 year round
  mutate(time_of_day = case_when(
    hour >= 6  & hour < 11 ~ "Morning",
    hour >= 11 & hour < 15 ~ "Midday",
    hour >= 15 & hour < 18 ~ "Afternoon",
    TRUE ~ "Night"
  )) %>%
  mutate(time_of_day = factor(time_of_day,
                              levels = c("Morning", "Midday", "Afternoon", "Night"))) %>%
  # select only the columns we need
  select(subfolder, original_image, crop_path, camera_trap_name,
         dt, date, hour, time_of_day, confidence)


summary(dfmetadata_clean)
```

Joining the labels onto the metadata

```{r}
# inner_join because a crop is only useful here if we both detected AND labelled it.
dfjoined <- dfclassifications_clean %>%
  inner_join(dfmetadata_clean, by = "crop_path")

# Sanity check: if this is 0 (or way below nrow(dfclassifications_clean)) then the
# paths aren't lining up and every plot below is silently wrong. Worth a look
# before trusting anything -- the 3 legacy rows labelled under the old flat
# folder layout are expected to drop out here.
nrow(dfclassifications_clean)
nrow(dfjoined)

# Just our two focal species
dffocal <- dfjoined %>%
  filter(species %in% c("Tortoise", "Iguana"))

dffocal %>% count(species)
```

## Location data (random-order sample)

`classifications.csv` is almost entirely SANTA FE 1001 (labelled in on-disk
folder order before random ordering existed), so any comparison across camera
locations built on it would mostly be measuring sampling order, not biology.
Camera-location comparisons below use `classifications_random_order.csv`
instead, via a parallel copy of the cleaning/joining steps above.

```{r}
dfclassifications_loc <- dfclassifications_random %>%
  clean_names() %>%
  rename_with(~ "species", ends_with("species")) %>%
  # Skip-button rows are written as "SKIPPED" rather than blank -- exclude
  # those along with blanks, same as the main classifications cleaning above.
  filter(species != "", species != "SKIPPED") %>%
  # If a crop was labelled out of the buffered folder it carries a _buf100 suffix that the metadata doesn't have. drop it so the two match.
  mutate(crop_path = str_replace(image, "_buf\\d+(\\.[^.]+)$", "\\1")) %>%
  mutate(crop_path = str_replace(crop_path, "^\\./", "")) %>%
  # This file mixes rows recorded relative to output/ ("cropped/SANTA FE .../x.JPG")
  # with rows relative to output/cropped/ ("SANTA FE .../x.JPG"); metadata's
  # crop_path is always the latter, so normalize by dropping a leading "cropped/".
  mutate(crop_path = str_replace(crop_path, "^cropped/", ""))

dfjoined_loc <- dfclassifications_loc %>%
  inner_join(dfmetadata_clean, by = "crop_path")

# Same sanity check as the main join above -- confirm paths actually matched.
nrow(dfclassifications_loc)
nrow(dfjoined_loc)

dffocal_loc <- dfjoined_loc %>%
  filter(species %in% c("Tortoise", "Iguana"))

dffocal_loc %>% count(species)
```

## Hypothesis 1: tortoises and iguanas are not in the same image

```{r}

dfphoto <- dfjoined %>%
  group_by(original_image, camera_trap_name, time_of_day, date) %>%
  summarize(
    has_tortoise = any(species == "Tortoise"),
    has_iguana   = any(species == "Iguana"),
    .groups = "drop"
  )

# For the plot we do want just the photos with at least one of our two species.
dfphoto_focal <- dfphoto %>%
  filter(has_tortoise | has_iguana) %>%
  mutate(composition = case_when(
    has_tortoise & has_iguana ~ "Both",
    has_tortoise ~ "Tortoise only",
    has_iguana ~ "Iguana only"
  )) %>%
  mutate(composition = factor(composition,
                              levels = c("Tortoise only", "Iguana only", "Both")))

dfphoto_focal %>% count(composition)
```

```{r}
dfphoto_focal %>%
  ggplot(aes(x = composition)) +
  geom_bar(fill = "#9a0101", alpha = 0.8) +
  labs(title = "Tortoises and Iguanas Rarely Share a Photo",
       x = "What was in the photo", y = "Number of photos") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

ggsave(filename = "../output/r_analysis/CoOccurrenceBarPlot.png", width = 10, height = 8, dpi = 300)
```

```{r}

dfphoto_focal %>%
  count(composition, name = "frequency") %>%
  mutate(percentage = round(frequency / sum(frequency) * 100)) %>%
  ggplot(aes(x = 2, y = frequency, fill = composition)) +
  geom_col(width = 1, color = "white") +
  # theta = "y" wraps the single stacked bar into a ring; xlim's lower bound
  # leaves the centre empty so it reads as a donut, not a filled pie.
  coord_polar(theta = "y") +
  xlim(0.5, 2.5) +
  geom_text(aes(label = paste0(percentage, "%")),
            position = position_stack(vjust = 0.5)) +
scale_fill_manual(values = c("#9a0101", "#ffcd01", "#808080")) +
  labs(title = "Tortoises and Iguanas Rarely Share a Photo",
       fill = "What was in the photo") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))

ggsave(filename = "../output/r_analysis/CoOccurrenceDonutPlot.png", width = 10, height = 8, dpi = 300)
```

```{r}
# "Rarely together" only means something against how often they'd land together
# by chance, so compare observed co-occurrence to independence across all photos.
cooccur_table <- table(tortoise = dfphoto$has_tortoise, iguana = dfphoto$has_iguana)
cooccur_table

chisq.test(cooccur_table)
```

## Hypothesis 2: they use the same camera at different times of day

```{r}
# The raw hour is more honest than the four bins -- the bins are our invention,
# this is what the cameras actually saw.
dffocal %>%
  ggplot(aes(x = hour, fill = species)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.5) +
  labs(title = "Hourly Detection Pattern by Species",
       x = "Hour of day", y = "Number of detections", fill = "Species") +
scale_fill_manual(values = c("#9a0101", "#ffcd01")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

ggsave(filename = "../output/r_analysis/HourlyActivityPlot.png", width = 10, height = 8, dpi = 300)
```

```{r}
dffocal_loc %>%
  count(camera_trap_name, time_of_day, species) %>%
  group_by(camera_trap_name, species) %>%
  # total = all detections of that species at that camera; proportion is the
  # share falling in each time-of-day bin. The bars are a proportion of N counts,
  # so the honest error bar is the standard error of a proportion: sqrt(p(1-p)/N).
  mutate(
    total      = sum(n),
    proportion = n / total,
    se         = sqrt(proportion * (1 - proportion) / total)
  ) %>%
  ungroup() %>%
  ggplot(aes(x = time_of_day, y = proportion, fill = species)) +
  geom_col(position = position_dodge(width = 0.9)) +
  # pmax keeps the lower whisker from dropping below 0, which a proportion can't be.
  geom_errorbar(aes(ymin = pmax(proportion - se, 0), ymax = proportion + se),
                position = position_dodge(width = 0.9), width = 0.2) +
scale_fill_manual(values = c("#9a0101", "#ffcd01")) +
  labs(title = "Tortoise and Iguana Activity Peaks at Different Times of Day",
       x = "Time of day", y = "Proportion of that species' detections",
       fill = "Species") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~camera_trap_name)

ggsave(filename = "../output/r_analysis/TimeOfDayByCameraPlot.png", width = 19, height = 8, dpi = 300)

```

T

```{r}
dffocal %>%
  count(camera_trap_name, time_of_day, species) %>%
  group_by(camera_trap_name, species) %>%
  mutate(proportion = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = time_of_day, y = proportion)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha = 0.3, color = "#991b1e") +
  labs(title = "Tortoise and Iguana Activity Peaks at Different Times of Day",
       x = "Time of day", y = "Proportion of that species' detections") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~species)

ggsave(filename = "../output/r_analysis/TimeOfDayBoxPlot.png", width = 10, height = 8, dpi = 300)
```

```{r}
# Is the time-of-day split different between the two species?
tod_table <- table(species = dffocal$species, time_of_day = dffocal$time_of_day)
tod_table

chisq.test(tod_table)
```

## Hypothesis 3: they are in different locations

```{r}
dffocal_loc %>%
  count(camera_trap_name, species, name = "frequency") %>%
  group_by(camera_trap_name) %>%
  mutate(proportion = frequency / sum(frequency)) %>%
  ungroup() %>%
  ggplot(aes(x = camera_trap_name, y = proportion, fill = species)) +
  geom_col(width = 0.9, color = "white") +
  # theta = "y" wraps each camera's bar into its own ring.
  coord_polar(theta = "y") +
  # camera_trap_name is discrete, so xlim() (a continuous scale) errors here.
  # Padding the low end of the discrete axis instead pushes the innermost ring
  # outward, which is what leaves the empty centre that makes it read as a donut.
  scale_x_discrete(expand = expansion(add = c(1.5, 0.6))) +
  labs(title = "Species Composition Differs Between Camera Locations",
       fill = "Species") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))

ggsave(filename = "../output/r_analysis/SpeciesByCameraDonutPlot.png", width = 10, height = 8, dpi = 300)
```

```{r}
dffocal_loc %>%
  count(species, camera_trap_name, name = "frequency") %>%
  ggplot(aes(y = frequency, axis1 = species, axis2 = camera_trap_name)) +
  geom_alluvium(aes(fill = species), width = 1/6) +
  geom_stratum(alpha = 0.5, width = 1/6) +
  geom_text(stat = "stratum", size = 3, aes(label = after_stat(stratum))) +
scale_fill_manual(values = c(
    "#9c0100",
    "#ffcd01"
  )) +
  scale_x_discrete(limits = c("Species", "Camera trap"), expand = c(0.1, 0.1)) +
  labs(title = "Which Camera Trap Each Species Was Detected",
       y = "Number of detections", fill = "Species") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

ggsave(filename = "../output/r_analysis/SpeciesByCameraAlluvialPlot.png", width = 10, height = 8, dpi = 300)
```

```{r}
dffocal_loc %>%
  count(species, camera_trap_name, name = "frequency") %>%
  PieDonut(aes(species, camera_trap_name, count = frequency),
           title = "Where Each Species Was Detected",
           r0 = 0.30, r1 = 0.9, color = "black", titlesize = 5)

# PieDonut draws straight to the graphics device rather than returning a plot
# object, so we can't hand it to ggsave -- record what was just drawn and replay
# it into a png, the same trick the iguana analysis uses.
recorded_plot <- recordPlot()
png(filename = "../output/r_analysis/SpeciesByCameraDonutNestedPlot.png", width = 10, height = 8, units = "in", res = 300)
replayPlot(recorded_plot)
dev.off()
```

```{r}
# Counts as well as proportions -- a camera with 3 detections and a camera with
# 3000 look identical once you divide, and that difference matters.
dffocal_loc %>%
  ggplot(aes(x = camera_trap_name, fill = species)) +
  geom_bar(position = "dodge") +
  labs(title = "Detections per Camera by Species",
       x = "Camera trap", y = "Number of detections", fill = "Species") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

ggsave(filename = "../output/r_analysis/DetectionsPerCameraPlot.png", width = 10, height = 8, dpi = 300)
```

```{r}
# Is species composition independent of camera?
cam_table <- table(species = dffocal_loc$species, camera = dffocal_loc$camera_trap_name)
cam_table

chisq.test(cam_table)
```

## Hypothesis 4: temporal avoidance is location-specific

Hypothesis 1 showed they're rarely in the same photo, Hypothesis 2 showed
their overall activity hours overlap, and Hypothesis 3 showed they favor
different cameras. The open question those leave: at a camera trap both
species actually use, do they still time-share it by shifting to different
parts of the day -- or is one species just dominant there around the clock?

A heatmap of species share per camera x time-of-day cell answers this more
directly than a faceted bar chart would: with 5 cameras x 4 time bins x 2
species, five separate donuts (one per camera) makes the cross-camera pattern
hard to compare at a glance, whereas a single grid puts every camera/time
combination in one view and highlights it with color.

```{r}
loc_time_species <- dffocal_loc %>%
  count(camera_trap_name, time_of_day, species) %>%
  group_by(camera_trap_name, time_of_day) %>%
  mutate(total = sum(n), iguana_share = sum(n[species == "Iguana"]) / total) %>%
  ungroup()

loc_time_species %>%
  distinct(camera_trap_name, time_of_day, total, iguana_share) %>%
  ggplot(aes(x = time_of_day, y = camera_trap_name, fill = iguana_share)) +
  geom_tile(color = "white") +
  # Label each cell with its N so an empty or near-empty cell (e.g. a camera
  # with one detection at 2am) doesn't read as strong evidence of anything.
  geom_text(aes(label = paste0(round(iguana_share * 100), "%\n(n=", total, ")")),
            size = 3, lineheight = 0.9) +
  scale_fill_gradient2(low = "#ffcd01", mid = "white", high = "#9a0101",
                        midpoint = 0.5, limits = c(0, 1), labels = scales::percent,
                        name = "Iguana share") +
  labs(title = "Species Composition by Camera and Time of Day",
       subtitle = "Deep red = Iguana-dominated, yellow = Tortoise-dominated, white = evenly split",
       x = "Time of day", y = "Camera trap") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))

ggsave(filename = "../output/r_analysis/TemporalAvoidanceHeatmap.png", width = 10, height = 6, dpi = 300)
```

If the two species were splitting time-of-day within a shared camera, that
camera's row should shift color across columns. If one species just owns that
camera outright, the row stays a flat color regardless of time of day.

```{r}
# Per-camera test: within each camera, is species composition independent of
# time of day? Small per-camera counts make the usual chi-square approximation
# unreliable, so simulate the p-value instead of trusting the asymptotic one.
dffocal_loc %>%
  group_by(camera_trap_name) %>%
  group_modify(~ {
    tbl <- table(species = .x$species, time_of_day = .x$time_of_day)
    test <- chisq.test(tbl, simulate.p.value = TRUE, B = 2000)
    tibble(n = sum(tbl), p_value = test$p.value)
  }) %>%
  ungroup()
```

Summary table

```{r}
species_stats <- dffocal %>%
  group_by(species, camera_trap_name) %>%
  summarize(
    detections   = n(),
    photos       = n_distinct(original_image),
    days         = n_distinct(date),
    first_seen   = min(date, na.rm = TRUE),
    last_seen    = max(date, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(species, camera_trap_name)

write.csv(species_stats, "../output/r_analysis/species_stats.csv", row.names = FALSE)

species_stats
```